Analyse gender in music psychology from journal publication data

### 2025-09-25 manuscript/manuscript.qmd
Worked on discussion: first and second paragraph (added literature/comparisons) and added limitations paragraph

Research across several scientific disciplines show that women tend to be less represented as authors compared to men [@banks2025women; @son2022scientific]. As the topics, approaches, and methods attract differential interest from men and women, it is important to recognise that fields of scholarship and academic disciplines vary considerably in terms of gender distribution [@huang2020]. For example, a large survey of authors by @gonzalez2020women showed that the hard sciences (N = 119,592) had the lowest prevalence of women (14.8%), whereas in the biological and social sciences (N = 262,122) the proportion was substantially higher (43.3%), but not equal. In psychology, which is perhaps the closest benchmark for the present focus on music psychology, female authors accounted for 45.2% of the sample (N = 90,067). Encouragingly, however, studies looking at various year ranges between 1960 until 2024 show a general trend that the ratio of women to men in authorships is improving [@gonzalez2020women; @ioannidis2023gender; @jemielniak2025gender; @sanchez2024analysis]. Yet, there is evidence that this improvement is plateauing [@jemielniak2025gender]. Further, although women are more likely to be first author, women are still generally unfairly underrepresented as the last author [@bruck2023bibliometric; @gonzalez2020women; @rock2021quantifying; @shah2021gender]. There could be several reasons for such a disparity: two key aspects that are typically explored in such **article analysis [TE: did you think citing specific studies here?]** papers: geographical locations and keyword/topics.

@@

Authorship is a crucial aspect in academia as a marker of scholarly success and strongly influences academic career progression, influencing recognition, funding, and advancement opportunities. However, there are several inequities within this aspect of academia, with women often being underrepresented in aspects such as authorships. The goal of the current paper was to provide a snapshot of authorship prevalence and roles across gender. While we recognise that gender is a complex and multidimensional social construct, and there are limitations to treating gender as a binary category, we used a binary male-female classification consistent with prior meta-science research on gender and authorship [@son2022scientific; @shah2021gender; @rock2021quantifying; @ni2021gendered; @gonzalez2020women; @wais2006]. As male-female authorship representations differ across different academic disciplines, it is worthwhile to diagnose the situation in music psychology.

In music psychology, the distribution of authors by gender remains unequal, with women representing 40.2% of authors. This proportion is higher than in musicology (34.8% @eerola2026inprep) and higher than in psychology more broadly, where women account for 33.2% of authors [@huang2020]. In terms of authorship roles, notable differences emerge. Women are relatively more likely than men to appear as first authors, a trend that has become characteristic of the field only in the past 15 years. This supports previous research showing that there is a positive trend with female representation increasing [@gonzalez2020women; @ioannidis2023gender; @jemielniak2025gender; @sanchez2024analysis]. In contrast, women remain underrepresented in the last authorship position, and this trend has not changed substantially over the last 15 years. This corresponds to the situation in other academic fields such as medicine [@bruck2023bibliometric; @shah2021gender], psychology [@gonzalez2020women], and biology [@rock2021quantifying]. Given that last authorship in the social sciences is not universally defined but is increasingly associated with seniority, supervisory responsibility, or principal investigator status [@drivas2024evolution], this underrepresentation suggests that women are less likely to occupy leadership roles such as lab heads or holders of substantial research resources. This pattern is consistent with broader evidence showing that women remain underrepresented in senior academic positions, including professorial appointments [@bakker2016tenure], invitations to contribute to journals, and authorship in high-prestige outlets [@holman2018gender]. Related analyses from science and medicine further show that women are more often associated with conducting experiments, whereas men are disproportionately represented in authorship roles linked to leadership and oversight [@macaluso2016science]. This suggests that the more 'leadership' author position is more likely to go to the male author.

@@

>>>>

Research across several scientific disciplines show that women tend to be less represented as authors compared to men [@banks2025women; @son2022scientific]. As the topics, approaches, and methods attract differential interest from men and women, it is important to recognise that fields of scholarship and academic disciplines vary considerably in terms of gender distribution [@huang2020]. For example, a large survey of authors by @gonzalez2020women showed that the hard sciences (N = 119,592) had the lowest prevalence of women (14.8%), whereas in the biological and social sciences (N = 262,122) the proportion was substantially higher (43.3%), but not equal. In psychology, which is perhaps the closest benchmark for the present focus on music psychology, female authors accounted for 45.2% of the sample (N = 90,067). Encouragingly, however, studies looking at various year ranges between 1960 until 2024 show a general trend that the ratio of women to men in authorships is improving [@gonzalez2020women; @ioannidis2023gender; @jemielniak2025gender; @sanchez2024analysis]. Yet, there is evidence that this improvement is plateauing [@jemielniak2025gender]. Further, although women are more likely to be first author, women are still generally unfairly underrepresented as the last author [@bruck2023bibliometric; @gonzalez2020women; @rock2021quantifying; @shah2021gender]. There could be several reasons for such a disparity. Two key aspects typically explored in bibliometric studies are geographical locations [e.g., @ioannidis2023gender] and keyword/topics [e.g., @bruck2023bibliometric].

@@

Authorship is a crucial aspect in academia as a marker of scholarly success and strongly influences academic career progression, influencing recognition, funding, and advancement opportunities. However, there are several inequities within this aspect of academia, with women often being underrepresented in authorships. As male-female authorship representations differ across different academic disciplines, it is worthwhile to diagnose the situation in the field of music psychology. The goal of the current paper was to provide a snapshot of authorship prevalence and roles across gender in music psychology journals, specifically Musicae Scientiae, Psychology of Music, Music Perception, Journal of New Music Research, and Music & Science. Overall in music psychology, we found that the distribution of authors by gender remains unequal, with women representing 40.2% of authors. This proportion is higher than in musicology (34.8% @eerola2026inprep) and higher than in psychology more broadly, where women account for a third of authors (33.2% [@huang2020]; 36.2% [@gonzalez2020women]). This representation is also much higher than in other STEM subjects of math, physics and computer (15% [@huang2020]) and physics topics like astronomy and bioscience (17.8%, @son2022scientific). However, overall women are still underrepresented. In exploring the nuances behind this, we further analysed differences between authorship role, number of citations, geography, and also keyword analysis.

In terms of authorship roles, notable differences emerge between authorship positions. Women are relatively more likely than men to appear as first authors, a trend that has become characteristic of the field only in the past 15 years. This supports previous research showing that there is a positive trend with female representation increasing [@gonzalez2020women; @ioannidis2023gender; @jemielniak2025gender; @sanchez2024analysis]. In contrast, women remain underrepresented in the last authorship position, and this trend has not changed substantially over the last 15 years. This corresponds to the situation in other academic fields such as medicine [@bruck2023bibliometric; @shah2021gender], psychology [@gonzalez2020women], and biology [@rock2021quantifying]. Given that last authorship in the social sciences is not universally defined but is increasingly associated with seniority, supervisory responsibility, or principal investigator status [@drivas2024evolution], this underrepresentation suggests that women are less likely to occupy leadership roles such as lab heads or holders of substantial research resources. This pattern is consistent with broader evidence showing that women remain underrepresented in senior academic positions, including professorial appointments [@bakker2016tenure], invitations to contribute to journals, and authorship in high-prestige outlets [@holman2018gender]. Related analyses from science and medicine further show that women are more often associated with conducting experiments, whereas men are disproportionately represented in authorship roles linked to leadership and oversight [@macaluso2016science]. This patterns is also consistent with the idea of a "leaky pipeline", where women are more likely to drop away from academic careers with increasing seniority, especially between docotral and (consolidated) researcher positions [@gonzalez2020women].

Additionally, women are less likely to self-promote accomplishments [@rudman1998self], and therefore might not push for the more senior role 
 This suggests that the more 'leadership' author position is more likely to go to the male author.

@@

There are certain limitations of the current research. First, the gender of the authors are not part of the database - we infer the gender based on the name using XYZ. Despite this, because the field is relatively small, a check from both authors meant that a certain number of the more ambiguous names (in terms of their gender) could be fixed (see accompanying code). Additionally, the error (confidence) of the gender attribution was so small, that it is unlikely that the remaining names that were not necessarily attributed correctly would significantly change the main conclusions. Another limitation is that while we recognise that gender is a complex and multidimensional social construct, there are limitations to treating gender as a binary category. We used a binary male-female classification consistent with prior meta-science research on gender and authorship [@son2022scientific; @shah2021gender; @rock2021quantifying; @ni2021gendered; @gonzalez2020women; @wais2006] and generally the number of non-binary authors is currently relatively low (0.8% @son2022scientific). Nonetheless, future research should also move towards using more inclusive, non-binary information.
